---
title: "HOSP Winter Storm"
author: "Danielle Ethier"
date: "2023-06-01"
output: html_document
editor_options: 
  chunk_output_type: console
---

Looking to see the effect of the 2022 Christmas Winter storm on HOSP. 

Load data and libraries

```{r load}

library(tidyverse)
require(lubridate)


data<-read_csv("Data/PFW_Canada_2022-2023.csv") 

```

Start by summarizing effort per week

```{r effort}

effort<-data %>% group_by(floor.week) %>% summarize(nefforthours = sum(effort_hrs_atleast, na.rm = TRUE), nlocation=n_distinct(loc_id)) 

ploteffort1<-ggplot(data=effort)+
  geom_point(aes(x=floor.week, y=nefforthours, colour = "red")) +
  geom_line(aes(x=floor.week, y=nefforthours, color = "red")) +
  geom_point(aes(x=floor.week, y=nlocation, colour = "blue"))+
  geom_line(aes(x=floor.week, y=nlocation, colour = "blue"))+
  theme_classic()

ploteffort1

#based on effort plot I will remove the last week of data collection.
#the data are therefore summarized between week 2022-11-12 to week 2023-04-01

data<-data %>% filter(floor.week!="2023-04-08")

```

Create events matrix to zero fill which will also include some covariates of interest

```{r events}

events<-data %>% select(Prov, Period, loc_id, year, month, day, date, snow_dep_atleast, housing_density) %>% distinct()

```

Select species of interest and explore data
```{r HOSP}

HOSP<-data %>% filter(species_code=="houspa")
#index for floor week ordered. 
HOSP$week <- as.integer(factor(HOSP$floor.week))

HOSP<-HOSP %>% group_by(Prov, floor.week, week) %>% summarise(n_loc=n_distinct(loc_id), tot_count=sum(how_many), mean_count=mean(how_many)) %>% filter(n_loc>10)

HOSP<-as.data.frame(HOSP)

ggplot(data=HOSP)+
  geom_point(aes(x=week, y=mean_count, colour=Prov))+
  facet_wrap(~Prov, scales="free")+
  geom_smooth(aes(x=week, y=mean_count, colour=Prov))+
  geom_vline(xintercept = 7)+
  theme_classic()

```

Mean count (zero filled) and percentage of feeders visited. Need to be seen at more than 10 locations, which removed the Atlantic Canada sites. 
```{r covariates}

HOSP<-data %>% filter(species_code=="houspa") %>% distinct()
HOSP$week <- as.integer(factor(HOSP$floor.week))

sp.data<-dplyr::left_join(events, HOSP, by=c("Prov", "Period", "date","loc_id", "day", "month", "year"), multiple = "all")
sp.data<-sp.data %>% filter(loc_id != "NA") 

# Zero fill the 'how_many' column
sp.data$how_many[is.na(sp.data$how_many)] <- 0 

sp.data <- sp.data %>% mutate(floor.week=floor_date(sp.data$date,unit="week")+6)

sp.data<- sp.data %>% group_by(Period) %>% mutate(week=as.integer(factor(floor.week)))

per.site.prov<-sp.data %>% group_by(Prov, floor.week, week) %>% 
mutate(loc_id_sp = ifelse(how_many==0, NA, loc_id)) %>%
summarize(tot.site=n_distinct(loc_id), n.site.sp=n_distinct(loc_id_sp, na.rm=TRUE), sp.sum=sum(how_many), mean.group=mean(how_many)) %>% filter(sp.sum > 0) %>% filter(tot.site>=10) %>% mutate(per.site=((n.site.sp)/tot.site)*100) %>% #subtract 1 for NAs
    dplyr::select(-sp.sum, -n.site.sp, -tot.site)

per.site.prov<-per.site.prov %>% filter(Prov %in% c("AB", "BC", "MB", "ON", "QC", "SK"))

ggplot(data=per.site.prov)+
  geom_point(aes(x=week, y=mean.group, colour=Prov))+
  facet_wrap(~Prov, scales="free")+
  geom_smooth(aes(x=week, y=mean.group, colour=Prov))+
  geom_vline(xintercept = 7)+
  theme_classic()

ggplot(data=per.site.prov)+
  geom_point(aes(x=week, y=per.site, colour=Prov))+
  facet_wrap(~Prov, scales="free")+
  geom_smooth(aes(x=week, y=per.site, colour=Prov))+
  geom_vline(xintercept = 7)+
  theme_classic()

```


Lets explore housing density to see if this makes a difference. 

```{r house}

sp.data<-NULL

HOSP<-data %>% filter(species_code=="houspa") %>% distinct()
HOSP$week <- as.integer(factor(HOSP$floor.week))
sp.data<-dplyr::left_join(events, HOSP, by=c("Prov", "Period", "loc_id", "day", "month", "year", "housing_density"), multiple = "all")
sp.data<-sp.data %>% filter(loc_id != "NA") 
sp.data<-sp.data %>% filter(housing_density != "NA") 
sp.data$housing_density<-as.factor(sp.data$housing_density)

#zero fill 'how_many", "floor.week", "week"
sp.data$how_many[is.na(sp.data$how_many)] <- 0 

sp.data <- sp.data %>% mutate(floor.week=floor_date(sp.data$date,unit="week")+6)

sp.data<- sp.data %>% group_by(Period) %>% mutate(week=as.integer(factor(floor.week)))

house.sum<-sp.data %>% group_by(Prov, floor.week, week, housing_density) %>% 
mutate(loc_id_sp = ifelse(how_many==0, NA, loc_id)) %>%
summarize(tot.site=n_distinct(loc_id), n.site.sp=n_distinct(loc_id_sp), sp.sum=sum(how_many), mean.group=mean(how_many)) %>% filter(sp.sum > 0) %>% filter(tot.site>=10) %>% mutate(per.site=((n.site.sp)/tot.site)*100) %>% #subtract 1 for NAs
    dplyr::select(-sp.sum, -n.site.sp, -tot.site)

ggplot(data=house.sum)+
  geom_point(aes(x=week, y=mean.group, colour=housing_density))+
  facet_wrap(~Prov, scales="free")+
  geom_smooth(aes(x=week, y=mean.group, colour=housing_density))+
  geom_vline(xintercept = 7)+
  theme_classic()

ggplot(data=house.sum)+
  geom_point(aes(x=week, y=per.site, colour=housing_density))+
  facet_wrap(~Prov, scales="free")+
 # geom_smooth(aes(x=week, y=mean.group, colour=housing_density))+
  geom_vline(xintercept = 7)+
  theme_classic()

```

##ASSESS MULTIPLE YEARS

Load data and libraries

```{r load}

library(tidyverse)

all<-read_csv("Data/PFW_Canada_All.csv") 

```

Start by summarizing effort per week

```{r effort}

effort<-all %>% group_by(floor.week) %>% summarize(nefforthours = sum(effort_hrs_atleast, na.rm = TRUE), nlocation=n_distinct(loc_id)) 

ploteffort1<-ggplot(data=effort)+
  geom_point(aes(x=floor.week, y=nefforthours, colour = "red")) +
  geom_line(aes(x=floor.week, y=nefforthours, color = "red")) +
  geom_point(aes(x=floor.week, y=nlocation, colour = "blue"))+
  geom_line(aes(x=floor.week, y=nlocation, colour = "blue"))+
  theme_classic()

ploteffort1

```

Create events matrix to zero fill which will also include some covariates of interest

```{r events}

events<-all %>% select(Prov, Period, loc_id, year, month, day, date) %>% distinct()

```

Select species of interest and explore data
```{r HOSP}

HOSP<-all %>% filter(species_code=="houspa")

#index for floor week ordered. 
HOSP<- HOSP %>% group_by(Period) %>% mutate(week=as.integer(factor(floor.week)))

HOSP<-HOSP %>% group_by(Prov, Period, floor.week, week) %>% summarise(n_loc=n_distinct(loc_id), tot_count=sum(how_many), mean_count=mean(how_many)) %>% filter(n_loc>10)

HOSP<-as.data.frame(HOSP)

ON<-HOSP %>% filter(Prov=="ON")

ggplot(data=ON)+
  geom_point(aes(x=week, y=mean_count))+
  facet_wrap(~Period, scales="free")+
  geom_smooth(aes(x=week, y=mean_count))+
  geom_vline(xintercept = 7)+
  theme_classic()

BC<-HOSP %>% filter(Prov=="BC")

ggplot(data=BC)+
  geom_point(aes(x=week, y=mean_count))+
  facet_wrap(~Period, scales="free")+
  geom_smooth(aes(x=week, y=mean_count))+
  geom_vline(xintercept = 7)+
  theme_classic()

```

Mean count (zero filled) and percentage of feeders visited. Need to be seen at more than 10 locations, which removed the Atlantic Canada sites. 
```{r covariates}

HOSP<-all %>% filter(species_code=="houspa") %>% distinct()

#index for floor week ordered. 
HOSP<- HOSP %>% group_by(Period) %>% mutate(week=as.integer(factor(floor.week)))

sp.data<-dplyr::left_join(events, HOSP, by=c("Prov", "Period", "loc_id", "day", "month", "year", "date"), multiple = "all")
sp.data<-sp.data %>% filter(loc_id != "NA") 

# Zero fill the 'how_many' column
sp.data$how_many[is.na(sp.data$how_many)] <- 0 

sp.data <- sp.data %>% mutate(floor.week=floor_date(sp.data$date,unit="week")+6)

sp.data<- sp.data %>% group_by(Period) %>% mutate(week=as.integer(factor(floor.week)))

#remove last floor.week
HOSP<-HOSP %>% filter(week <=20)

per.site.prov<-sp.data %>% group_by(Prov, Period, floor.week, week) %>% 
mutate(loc_id_sp = ifelse(how_many==0, NA, loc_id)) %>%
summarize(tot.site=n_distinct(loc_id), n.site.sp=n_distinct(loc_id_sp, na.rm=TRUE), sp.sum=sum(how_many), mean.group=mean(how_many)) %>% 
  #filter(sp.sum > 0) %>% filter(tot.site>=10) %>%
  mutate(per.site=((n.site.sp)/tot.site)*100) %>% #subtract 1 for NAs
    dplyr::select(-sp.sum, -n.site.sp, -tot.site)

ON.per.site<-per.site.prov %>% filter(Prov=="ON")

ggplot(data=ON.per.site)+
  geom_point(aes(x=week, y=mean.group))+
  facet_wrap(~Period, scales="free")+
  geom_smooth(aes(x=week, y=mean.group))+
  geom_vline(xintercept = 7)+
  theme_classic()

ggplot(data=ON.per.site)+
  geom_point(aes(x=week, y=per.site))+
  facet_wrap(~Period, scales="free")+
  geom_smooth(aes(x=week, y=per.site))+
  geom_vline(xintercept = 7)+
  theme_classic()

BC.per.site<-per.site.prov %>% filter(Prov=="ON")

ggplot(data=BC.per.site)+
  geom_point(aes(x=week, y=mean.group))+
  facet_wrap(~Period, scales="free")+
  geom_smooth(aes(x=week, y=mean.group))+
  geom_vline(xintercept = 7)+
  theme_classic()

ggplot(data=BC.per.site)+
  geom_point(aes(x=week, y=per.site))+
  facet_wrap(~Period, scales="free")+
  geom_smooth(aes(x=week, y=per.site))+
  geom_vline(xintercept = 7)+
  theme_classic()

```


