---
title: "PFWPlots_AnnualReport"
author: "Danielle Ethier"
date: "11/01/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

This script is not part of the PFW R user guide. It was created to automate graphs for Kerrie's reports. This script, however, uses outputs from the PFW user guide, found in the `Outputs` folder. 

Start by loading packages and the 
```{r load packages}

require(tidyverse)
require(ggplot2)
require("gridExtra")

out.dir <- paste("Output/")
dat.dir <- paste("Data/")
plot.dir<- paste("Plot/")

```

Load data from `Output` folder. 
```{r data}

regionP<-read.csv("Output/ % sites with species at least once in season and mean group size_region.csv")
  
provP<-read.csv("Output/ % sites with species at least once in season and mean group size_prov.csv")
  
natP<-read.csv("Output/ % sites with species at least once in season and mean group size_nat.csv")


```

Add column to the national summary and append it to the regional and provincial datatables.
```{r nat}

nat_Pp<-add_column(natP, Prov="Canada", .after=0)
provP<-rbind(provP, nat_Pp)
provP<-provP %>% filter(Species != "NA")
# Relevel group factors 
provP$Prov <- factor(provP$Prov,                 
                         levels = c("BC", "AB", "SK", "MB", "ON", "QC", "NB", "NS", "PE", "NL", "North", "Canada"))


nat_Pr<-add_column(natP, Region="Canada", .after=0)
regionP<-rbind(regionP, nat_Pr)
regionP<-regionP %>% filter(Species!="NA")
# Relevel group factors 
regionP$Region <- factor(regionP$Region,                 
                         levels = c("BC", "PR", "ON", "QC", "AT", "Canada"))


```

Add common names from plotting and taxon order

```{r names}

sp<-read.csv("Data/PFW_species_codes.csv")
sp<-sp %>% dplyr::select(REPORT_AS, PRIMARY_COM_NAME, TAXON_ORDER) %>% distinct(REPORT_AS, .keep_all=TRUE)

provP<-left_join(provP, sp, by = c("Species"="REPORT_AS"))
provP<-provP %>% arrange(TAXON_ORDER)
provP$Period<-as.numeric(provP$Period)

regionP<-left_join(regionP, sp, by = c("Species"="REPORT_AS"))
regionP<-regionP %>% arrange(TAXON_ORDER)
regionP$Period<-as.numeric(regionP$Period)


```

Prepare list of species for loop and create plots. 

```{r species}

##Regional Summary
sp.list<-unique(regionP$PRIMARY_COM_NAME)

#Regional species loop 
for(n in 1:length(sp.list)) {
  
   # n<-1 #for testing each species

  #Provincial Zero-fill  
  sp.data1 <-NULL #clear previous dataframe
  sp.data1 <- filter(regionP, PRIMARY_COM_NAME == sp.list[n]) %>%
      droplevels()
  sp.data1<-sp.data1 %>% drop_na(Region)
  
if(nrow(sp.data1)>40){ #only continue if there is sufficient data
  
pdf(paste(plot.dir, sp.list[n], "_ProvZeroFill",".pdf", sep=""), onefile = TRUE, height = 10, width = 8, paper = "letter")

  # Plot the mean number
top.plot1<-ggplot(sp.data1, aes(Period, MeanGroup))+
  geom_line(aes(colour=Region), size=1, show.legend = FALSE)+
  theme_classic()+
  theme(text=element_text(size=16))+
  theme(axis.text.x = element_text(angle = 90))+
  ylab("Mean number per site")+
  xlab("")+
  ggtitle(sp.list[n]) 

# Plot the % feeders 
bottom.plot1<-ggplot(sp.data1, aes(Period, PercentSite))+
  geom_line(aes(colour=Region), size=1)+
  theme_classic()+
  xlab("Period")+
  theme(axis.text.x = element_text(angle = 90))+
  theme(text=element_text(size=16))+
  ylab("% feeder sites")+
  theme(legend.position = "bottom")+
  theme(legend.title = element_blank())+
  guides(colour = guide_legend(nrow = 1))
 

grid.arrange(top.plot1, bottom.plot1)

while(!is.null(dev.list())) dev.off()

} #end if nrow
    } #end species loop
```

Now plot the number of site as a means of reporting effort

```{r effort}

#Load the events dataset
event<-read.csv("Output/Events.csv")
range<-read.csv("Output/Range.csv")

blk<-range %>% dplyr::select(loc_id, Prov, region, BlockCode) %>% distinct(loc_id, .keep_all = TRUE)

eff<-left_join(event, blk, by="loc_id")

## Number of sites monitored
loc_id_nat<-eff %>% group_by(Period) %>% summarize (n_loc = n_distinct(loc_id)) %>% drop_na()

pdf(paste(plot.dir, "Number count sites Canada.pdf", sep=""), onefile = TRUE)

ggplot(loc_id_nat, aes(x=Period, y=n_loc))+
  geom_line(size=1.5)+
  theme_classic()+
  theme(text=element_text(size=16))+
  ylab("Number of count sites")+
  geom_point(aes(x=2020, y=2172), colour="red", size=5)+
  annotate("text", x = 2017, y = 2200, label = "COVID-19", colour="red", size = 5)

while(!is.null(dev.list())) dev.off()

```
